\input{mintmod.tex}
\MPragma{MathSkip}

%\Mtikzexternalize

\begin{document}
%\include{vbkm03_eng}
%\include{vbkm02_eng}
\MSection{Inequalities in one Variable}
\MLabel{VBKM03}
\MSetSectionID{ungl}

\begin{MSectionStart}
\MDeclareSiteUXID{VBKM03_START}

Inequalities arise by relating terms using one of the comparing symbols $\leq$, $<$, $\geq$, or $>$. Simple 
inequalities usually have intervals as their solution sets. But the solution of inequalities is often
more difficult than the solution of equations. Hence, specific types of inequalities will be explained
in more detail.

This module consists of:

\begin{itemize}
\item{Section~\MNRef{M03_Ungleichungen}: \MSRef{M03_Ungleichungen}{Inequalities and their Solution Sets},}
\item{Section~\MNRef{M03_Umformen}: \MSRef{M03_Umformen}{Transformation of Inequalities},}
\item{Section~\MNRef{M03_Betragsungleichungen}: \MSRef{M03_Betragsungleichungen}{Absolute Value Inequalities and 
Quadratic Inequalities},}
\item{and Section~\MNRef{M03_Abschlusstest}: \MSRef{M03_Abschlusstest}{Final Test}.}
\end{itemize}

\end{MSectionStart}

\MSubsection{Inequalities and their Solution Sets}
\MLabel{M03_Ungleichungen}

\begin{MIntro}

\begin{MInfo}
\MDeclareSiteUXID{VBKM03_UngleichungenIntro}
If two numbers are related by one of the \MEntry{comparing symbols}{comparing symbols} 
$\leq$, $<$, $\geq$, or $>$, a statement is generated that can be true or false depending on 
the numbers:
\begin{itemize}
\item{$a<b$ (reads: ``$a$ is strictly less than $b$'' or simply ``$a$ is less than $b$'') is true if the number $a$ is less than and not equal to $b$.}
\item{$a \leq b$ (reads: ``$a$ is less than $b$'') is true if $a$ is less than or equal to $b$.}
\item{$a>b$ (reads: ``$a$ is strictly greater than $b$'' or simply ``$a$ is greater than $b$'') is true if the number $a$ is greater and not equal to $b$.}
\item{$a \geq b$ (reads: ``$a$ is greater than $b$'') is true if the number $a$ is greater than or equal to $b$.}
\end{itemize}
\end{MInfo}

The relating symbols describe how the given values are related to each other on the number line: 
$a<b$ means that $a$ is to the left of $b$ on the number line.

\begin{MExample}
The statements $2<4$, $-12\leq 2$, $4>1$, and $3\geq 3$ are true,
but the statements $2<\sqrt2$ and $3>3$ are false.

\begin{center}
\MTikzAuto{%
\begin{tikzpicture}
% reelle Achse
\draw[->,color=black] (-1,0.0) -- (5,0.0);
\foreach \x in {-1, 0, 1, 2, 3, 4}
\draw[shift={(\x,0)},color=black] (0pt,2pt) -- (0pt,-2pt) node[below] {\footnotesize $\x$};
\draw (4.9,-0.3) node[] {$\mathbb{R}$};
% Enden:
\draw [fill = blue] (2,0) circle (1.5pt);
\draw [fill = blue] (4,0) circle (1.5pt);
\end{tikzpicture}
}%

On the number line, the number $2$ is to the left of the number $4$, thus $2<4$.
\end{center}

\end{MExample}

Here, $a<b$ means the same as $b>a$, likewise $a\leq b$ means the same as $b\geq a$. But it
should be noted that the opposite of the statement $a<b$ is the statement $a\geq b$ and not
$a>b$. If terms with a variable occur in an inequality, the problem is to find the number range
of the variable such that the inequality is true. 
\end{MIntro}

\begin{MXContent}{Solving simple Inequalities}{Solving}{STD}
\MDeclareSiteUXID{VBKM03_EinfacheUngleichungen}

If the variable occurs isolated in the inequality, the solution set is an interval, see also info
box \MRef{VBKM01_Intervalle}: 


\begin{MInfo}
\MLabel{M03_Aufloesungen}
The \MEntry{solved inequalities}{inequalities (solved)} 
have the following \MEntry{intervals}{intervals} as their solution sets:
\begin{itemize}
\item{$x< a$ has the solution set $\MoIl[\left] -\infty\MIntvlSep a\MoIr[\right]$, i.e.\ all $x$ less than $a$.}
\item{$x\leq a$ has the solution set $\MoIl[\left] -\infty\MIntvlSep a\right]$, 
i.e. all $x$ less than or equal to $a$.}
\item{$x> a$ has the solution set $\MoIl[\left] a\MIntvlSep \infty\MoIr[\right]$, i.e.\ all
 $x$ greater than $a$.}
\item{$x\geq a$ has the solution set $\left[a\MIntvlSep \infty\MoIr[\right]$, i.e.\ all $x$
greater than or equal to $a$.}
\end{itemize}
Here, $x$ is the variable and $a$ is a specific value. 
If the variable does not occur in the inequality anymore, the solution set is either
$\R=\MoIl[\left] -\infty\MIntvlSep \infty\MoIr[\right]$ if the inequality is satisfied, 
or the empty set $\lbrace \rbrace$ if the inequality is not satisfied.
\end{MInfo}

The symbol $\infty$ means \MEntry{infinity}{infinity}. A finite interval has the form 
$\MoIl a\MIntvlSep b\MoIr$ which reads ``all numbers between $a$ and $b$''. If the interval
is to be finite only on one side, the other interval boundary can be replaced by the symbol 
$\infty$ (right-hand side) or $-\infty$ (left-hand side).

As for equations one tries to find a solved inequality by applying transformations that do
not change the solution set. From the solved inequality the solution set can be easily seen.


\begin{MInfo}
\MLabel{VBKM03_AequivalenzumformungenUngleichungen}
To obtain a solved inequality from an unsolved inequality the following 
\MEntry{equivalent transformations}{equivalent transformations (inequality)} are allowed:
\begin{itemize}
\item{adding a constant to both sides of the inequality: $a<b$ is equivalent to $a+c<b+c$.}
\item{multiplying both sides of the inequality by a positive constant: $a<b$ 
is equivalent to $a\cdot c<b\cdot c$ if $c>0$.}
\item{multiplying both sides of the inequality by a negative constant and inverting the 
comparing symbol: $a<b$ is equivalent to $a\cdot c>b\cdot c$ if $c<0$.}
\end{itemize}
\end{MInfo}

\begin{MExample}
The inequality $-\frac34x-\frac12<2$ is solved stepwise by the above transformations:
\begin{eqnarray*}
&&-\frac34x-\frac12 < 2 \;\; \MSep +\frac12\ \\
&\Leftrightarrow&-\frac34x < 2+\frac12 \;\; \MSep \cdot\left({-\frac43}\right)\ \\
&\Leftrightarrow&x > -\frac43\left({2+\frac12}\right) \;\; \MSep \;\text{simplifying}\\
&\Leftrightarrow&x >  -\frac{20}{6} \;=\; -\frac{10}{3} \MDFPeriod
\end{eqnarray*}
So, the initial inequality has the solution set 
 $\MoIl[\left] -\frac{10}{3}\MIntvlSep \infty\MoIr[\right]$. 
Importantly, multiplying the inequality by the negative number $-\frac43$ inverts the 
comparing symbol.
\end{MExample}

\begin{MExercise}
Are the following inequalities true or false?

\begin{MQuestionGroup}
\begin{tabular}{lll}
\MLCheckbox{0}{UG1} & \ \ &  $\frac12>1-\frac13$\\
\MLCheckbox{1}{UG2} & \ \ & $a^2\geq 2a b-b^2$ (where $a$ and $b$ are unknown numbers)\\
\MLCheckbox{1}{UG3} & \ \ & $\frac12<\frac23<\frac34$\\
\MLCheckbox{0}{UG4} & \ \ & Let $a<b$, then also $a^2<b^2$.
\end{tabular}
\end{MQuestionGroup}
\MGroupButton{Check input}

\begin{MHint}{Solution}
The first inequality can be simplified to $\frac12>\frac23$, which, after multiplying by $6$, 
is equivalent to $3>4$. This statement is false. The second inequality can be simplified by 
collecting all numbers on the left-hand side: $a^2-2a b+b^2\geq 0$. Since $a^2-2a b+b^2=(a-b)^2$,
this statement is true for all $a$ and $b$. Multiplying the third chain of inequalities by the
least common denominator $12$ results in the chain of inequalities $6<8<9$. This statement is true.
In contrast, the last statement is false, since for example, for $a=-1$ and $b=1$, the term
$a^2=1$ is not less than $b^2=1$. Taking the square of terms is not an equivalent transformation.
\end{MHint}
\end{MExercise}


\begin{MExercise}
Find the solution sets of the following inequalities.
\begin{MExerciseItems}
\item{$2x+1> 3x-1$ has the solution interval \MEquationItem{$\ML$}{\MLIntervalQuestion{30}{(-infty,2)}{5}{TXH1}}.}
\item{$-3x-\frac12\leq x+\frac12$ has the solution interval \MEquationItem{$\ML$}{\MLIntervalQuestion{30}{[-1/4,infty)}{5}{TXH2}}.}
\item{$x-\frac12\leq x+\frac12$ has the solution interval \MEquationItem{$\ML$}{\MLIntervalQuestion{30}{(-infty,infty)}{5}{TXH3}}.}
\end{MExerciseItems}
\MInputHint{Enter the intervals in the form \texttt{(a;b)}, for the interval boundaries also fractions and
\texttt{infinity} or \texttt{-infinity} can be used. Take care whether the interval boundaries are included 
or excluded.}

\begin{MHint}{Solution}
Transformation of the first inequality results in
\begin{eqnarray*}
&& 2x+1 > 3x-1\;\; \MSep+1\ \\
&\Leftrightarrow& 2x+2 > 3x\;\; \MSep-2x\ \\
&\Leftrightarrow&2 > x
\end{eqnarray*}
and hence the solution interval is $\ML=\MoIl[\left] -\infty\MIntvlSep 2\MoIr[\right]$. 
Transformation of the second inequality results in
\begin{eqnarray*}
&&-3x-\frac12\leq x+\frac12 \;\; \MSep +3x-\frac12\ \\
&\Leftrightarrow&-1\leq 4x \;\; \MSep \cdot \frac14\ \\
&\Leftrightarrow&-\frac14\leq  x
\end{eqnarray*}
and hence $\ML=\left[-\frac14\MIntvlSep \infty\MoIr[\right]$. 
Transformation of the third inequality results in
\begin{eqnarray*}
&&x-\frac12\leq x+\frac12\;\; \MSep-x\ \\
&\Leftrightarrow&-\frac12\leq \frac12 \MDFPeriod
\end{eqnarray*}
This statement does not depend on $x\in\R$ and is always true, 
thus the solution set is $\ML=\R=\MoIl[\left] -\infty\MIntvlSep \infty\MoIr[\right]$.
\end{MHint}
\end{MExercise}

\begin{MInfo}
An inequality in one variable $x$ is \MEntry{linear}{inequality (linear)} if on both sides of the 
inequality only multiples of $x$ and constants occur. Each linear inequality can be transformed 
into a solved inequality by one of the equivalent transformations described in the info box
\MRef{M03_Aufloesungen}.
\end{MInfo}

\end{MXContent}

\begin{MXContent}{Specific Transformations}{Specific Transformations}{STD}
\MDeclareSiteUXID{VBKM03_SpezielleUmformungen}
The following equivalent transformations are useful if the variable occurs 
in the denominator of an expression. But they can only be applied under certain
restrictions:

\begin{MInfo}
Under the restriction that none of the occurring denominators is zero (the corresponding variable values are
 never solutions) and the fractions on both sides have the same sign, the reciprocal can be taken
on both sides of the inequality while inverting the comparing symbol.
\end{MInfo}

\begin{MExample}
For example, the inequality $\frac1{2x}\leq \frac1{3x}$ is equivalent to $2x\geq 3x$
(comparing symbol inverted) as long as $x\not=0$. The new inequality has the solution
set $\MoIl[\left] -\infty\MIntvlSep 0\right]$. However, since the value $x=0$ was excluded (and 
does not belong to the domain of the initial inequality either) the solution set of 
$\frac1{2x}\leq \frac1{3x}$ is $\ML=\MoIl[\left] -\infty\MIntvlSep 0\MoIr[\right]$.
\end{MExample}

\begin{MExercise}
Find the solution sets of the following inequalities.
\begin{MExerciseItems}
\item{$\frac1x\geq\frac13$ has the solution set \MEquationItem{$\ML$}{\MLIntervalQuestion{20}{(0;3]}{3}{KKL1}}.}
\item{$\frac1x<\frac1{\sqrt{x}}$ has the solution set \MEquationItem{$\ML$}{\MLIntervalQuestion{20}{(1;infty)}{3}{KKL2}}.}
\end{MExerciseItems}

\begin{MHint}{Solution}
For the first inequality, the value $x=0$ is not in the domain, hence this value is excluded. For 
$x>0$, taking the reciprocal while inverting the comparing symbol is allowed and results in
$x\leq 3$. Together with the condition above the solution interval is $\ML=\MoIl 0\MIntvlSep 3]$.
For $x<0$ the reciprocal rule cannot be applied. However, it can be seen, even without any rule, that
none of the values $x<0$ can be a solution, since then $\frac1x$ is negative as well and not greater than
or equal to $\frac13$.

The domain of the second inequality is $\MoIl 0\MIntvlSep \infty\MoIr$, since 
only for these values of $x$ taking the square root is
defined and only for $x\neq 0$ the denominators are non-zero. On the domain, 
taking the reciprocal while inverting the comparing symbol is 
allowed and results in $x>\sqrt{x}$. Since $\sqrt{x}>0$, the inequality can be 
divided by $\sqrt{x}$ resulting in $\sqrt{x}>1$. This inequality has the solution
set $\ML=\MoIl 1\MIntvlSep \infty\MoIr$ which occurs also in the domain.
\end{MHint}

\end{MExercise}

Please note for the last part of the exercise:

\begin{MInfo}
Taking the square on both sides of an inequality is not an equivalent transformation and 
possibly does change the solution set.
\end{MInfo}

For example, $x=-2$ is no solution of $x>\sqrt{x}$, but indeed a solution of $x^2>x$. However,
this transformation can be applied if the case analysis for the transformation
is carried out correctly and the domain of the initial inequality is taken into account. This method is
described in more detail in the next section.
\end{MXContent}

\MSubsection{Transformation of Inequalities}
\MLabel{M03_Umformen}

\begin{MXContent}{Transformation with Case Analysis}{Case Analysis}{STD}
\MDeclareSiteUXID{VBKM03_UmformungenFallunterscheidungen}
The simple linear transformations described in the previous section are equivalent transformations.
They do not change the solution set of the corresponding inequality. For nonlinear
inequalities advanced solution methods are required. Usually, these methods need 
a case analysis depending on the sign, since, in contrast to the situation for
equations described in Modul~\MNRef{VBKM02}, now also the inequality can be inverted during 
transformation.

 
\begin{MInfo}
If an inequality is multiplied by a term in which the variable $x$ occurs, a case analysis 
is required and for each case the transformation has to be considered separately: 

\begin{itemize}
\item{For those values of $x$, for which the multiplied term is positive, the comparing symbol
of the inequality is unchanged.}
\item{For those values of $x$, for which the multiplied term is negative, the comparing symbol
of the inequality is inverted.}
\item{The case that the multiplied term is zero has to be excluded during the transformation and
has to be considered separately, if necessary.}
\end{itemize}
\ \\ \ \\
The solution sets found in the individual cases have to be checked with respect to the case conditions
as described for the solution of \MSRef{VBKM02_FallBetrag}{absolute value equations}.
\end{MInfo}

In contrast, adding terms in which the variable occurs, does not require a case analysis. Usually, 
transformations involving case analyses are mandatory if the variable occurs in the denominator or 
in a composite term.


\begin{MExample}
The inequality $\frac1{2x}\leq 1$ can be simplified by multiplying both sides of the inequality
by the term $2x$:

\begin{itemize}
\item{Under the condition $x>0$ this results in the new inequality $1\leq 2x$. It has the solution set
$\ML_1=\left[\frac12\MIntvlSep \infty\MoIr[\right]$. The condition $x>0$ is satisfied by all elements of the solution 
set.}
\item{Under the condition $x<0$ this results in the new inequality $1\geq 2x$. It has the solution set
 $\MoIl[\left] -\infty\MIntvlSep \frac12\right]$. Because of the additional
condition $x<0$ only the elements of the set 
$\ML_2=\MoIl[\left] -\infty\MIntvlSep 0\MoIr[\right]$ are solutions.}
\item{The single case $x=0$ is no solution since this value is not in the domain of the inequality. 
In this case multiplying the inequality by $x$ is not allowed.}
\end{itemize}
So, altogether one obtains the union set 
$\ML=\ML_1\cup \ML_2=\R\MSetminus\left[0\MIntvlSep \frac12\MoIr[\right]$ as solution set:
\ \\ \ \\
\begin{center}
\MTikzAuto{%
\begin{tikzpicture}
% reelle Achse
\draw[->,color=black] (-1,0.0) -- (5,0.0);
\foreach \x in {-1, 0, 1, 2, 3, 4}
\draw[shift={(\x,0)},color=black] (0pt,2pt) -- (0pt,-2pt) node[below] {\footnotesize $\x$};
\draw (4.9,-0.3) node[] {$\mathbb{R}$};
% Enden:
\draw [line width=2.0pt,color=blue] (-1,0.0)-- (0,0.0);
\draw [line width=2.0pt,color=blue] (0.5,0.0)-- (5,0.0);
\draw [fill = blue] (0.5,0) circle (1.5pt);
\draw [fill = white] (0,0) circle (1.5pt);
\end{tikzpicture}
}
\end{center}
\end{MExample}

As in Modul~\MNRef{VBKM02} the following statement holds for the solution set.

\begin{MInfo}
The cases have to be chosen such that all elements of the domain of the inequality are covered. 
For the solution set in an individual case, it has to be checked that the solution set satisfies the 
corresponding case condition. For any case, the resulting solution set has to be reduced to
the solution subset satisfying the case condition. The union of the solution sets for the individual cases
is the solution set of the initial inequality.
\end{MInfo}

\end{MXContent}

\begin{MExercises}
\MDeclareSiteUXID{VBKM03_Fallunterscheidungen_Exercises}
If the inequality is multiplied by a composite term, it has to be investigated precisely for which values
of $x$ the case analysis has to be done:

\begin{MExercise}
Find the solution set of the inequality $\frac1{4-2x}<3$. 
The domain of the inequality is $D=\R\MSetminus \lbrace 2\rbrace$ since only for these
values of $x$ the denominator is non-zero. If the inequality is multiplied by the term
$4-2x$, three cases have to be distinguished. Fill in the blanks in the following text 
accordingly:

\begin{MExerciseItems}
\item{On the interval \MLIntervalQuestion{15}{(-infty,2)}{4}{GOM1} the term is positive, the comparing symbol 
remains unchanged, and the new inequality reads $1\:<\:$\MLSimplifyQuestion{15}{3*(4-2*x)}{5}{x}{5}{0}{SIMPLE2}.
Linear transformations result in the solution set 
\MEquationItem{$\ML_1$}{\MLIntervalQuestion{20}{(-infty,11/6)}{3}{MIXGOM}}. 
The elements of this set satisfy the case condition.}
\item{On the interval \MLIntervalQuestion{15}{(2,infty)}{4}{GOM2} the term is negative, 
the comparing symbol is inverted. Initially, the new inequality has the solution set 
\MLIntervalQuestion{20}{(11/6,infty)}{4}{INT1}, because of the case condition only the 
subset \MEquationItem{$\ML_2$}{\MLIntervalQuestion{20}{(2,infty)}{4}{GOM3}} is allowed.}
\item{The single value $x=2$ is no solution of the initial inequality since 
it is not in \MLQuestion{25}{domain}{UGX}.}
\end{MExerciseItems}

Sketch the solution set of the inequality and indicate the boundary points.

\begin{MHint}{Solution}
On the interval $\MoIl[\left] -\infty\MIntvlSep 2\MoIr[\right]$ the term is positive. 
The corresponding solution set is $\MoIl[\left] -\infty\MIntvlSep \frac{11}{6}\MoIr[\right]$.
In contrast, on the interval $\MoIl 2\MIntvlSep \infty\MoIr$ the term is negative, 
the comparing symbol is inverted. Initially, the new inequality has the solution set
 $\MoIl[\left] \frac{11}{6}\MIntvlSep \infty\MoIr[\right]$, because of the case condition
$x>2$ only the subset $\ML_2=\MoIl 2\MIntvlSep \infty\MoIr$ is allowed.
So, altogether the union set $\ML=\ML_1\cup\ML_2=\R\MSetminus \left[\frac{11}6\MIntvlSep 2\right]$
is the solution set of the initial inequality excluding the boundary points:

\begin{center}
\MTikzAuto{%
\begin{tikzpicture}
% reelle Achse
\draw[->,color=black] (-1,0.0) -- (5,0.0);
\foreach \x in {-1, 0, 1, 2, 3, 4}
\draw[shift={(\x,0)},color=black] (0pt,2pt) -- (0pt,-2pt) node[below] {\footnotesize $\x$};
\draw (4.9,-0.3) node[] {$\mathbb{R}$};
% Enden:
\draw [line width=2.0pt,color=blue] (-1,0.0)-- (1.83,0.0);
\draw [line width=2.0pt,color=blue] (2,0.0)-- (5,0.0);
\draw [fill = white] (1.83,0) circle (1.5pt);
\draw [fill = white] (2,0) circle (1.5pt);
\end{tikzpicture}
}
\end{center}
\end{MHint}
\end{MExercise}

\begin{MExercise}
The solution set of the inequality $\frac{x-1}{x-2}\leq 1$ is 
\MEquationItem{$\ML$}{\MLIntervalQuestion{20}{(-infty,2)}{4}{IGU1}}.

\begin{MHint}{Solution}
The domain of the inequality is $D=\R\MSetminus\lbrace 2\rbrace$.

\begin{itemize}
\item{For $x>2$, multiplying the inequality by the term $x-2$ results in $x-1\leq x-2$, 
which is equivalent to the false statement $-1\leq -2$. Thus, this case does not contribute a solution to the solution set.}
\item{For $x<2$, multiplying the inequality by the term $x-2$ results in $x-1\geq x-2$, 
which is equivalent to the true statement $-1\geq -2$. 
Because of the case condition the solution interval for this case is only 
$\ML_2=\MoIl[\left] -\infty\MIntvlSep 2\MoIr[\right]$.}
\item{The single value $x=2$ is no solution.}
\end{itemize}
So, altogether the solution set is 
 $\ML=\MoIl[\left] -\infty\MIntvlSep 2\MoIr[\right]$ 
excluding the boundary points (even though the comparing symbol $\leq$ occurred in the initial
inequality).
\end{MHint}
\end{MExercise}


\begin{MExercise}
The solution set of the inequality $\frac1{1-\sqrt{x}}<1+\sqrt{x}$ is \MEquationItem{$\ML$}{\MLIntervalQuestion{20}{(1,infty)}{4}{IGU2}}.
\ \\ \ \\
\begin{MHint}{Solution}
The domain of the inequality is $D=[0\MIntvlSep \infty\MoIr\MSetminus \lbrace 1\rbrace$
since only for these values of $x$ the square root is defined and the denominator is non-zero.
\begin{itemize}
\item{For $0\leq x<1$, multiplying the inequality by the term $1-\sqrt{x}$ results in
  $1<(1+\sqrt{x})(1-\sqrt{x})$, which is equivalent to $1<1-x$. This 
inequality is satisfied for $x<0$, but these values of $x$ violate the case condition and 
thus, they are not in the solution set.}
\item{For $x>1$, multiplying the inequality by the term $1-\sqrt{x}$ results in $1>1-x$, 
which is equivalent to $x>0$. But only the values of $x$ in the interval
  $\MoIl 1\MIntvlSep \infty\MoIr$ satisfy the case condition, hence $\ML=\MoIl 1\MIntvlSep \infty\MoIr$ 
is the only solution interval of the initial inequality.}
\item{The single value $x=1$ is no solution.}
\end{itemize}
\end{MHint}
\end{MExercise}


\end{MExercises}

\MSubsection{Absolute Value Inequalities and Quadratic Inequalities}
\MLabel{M03_Betragsungleichungen}

\begin{MIntro}
\MDeclareSiteUXID{VBKM03_Betragsungleichungen_Intro}
As in the approach in Modul~\MNRef{VBKM02} and in the previous section 
\MEntry{absolute values}{inequalities (absolute values)} in inequalities are solved 
by a case analysis:

\begin{MInfo}
To solve an \MEntry{absolute value inequality}{absolute value inequality} two cases are distinguished:

\begin{itemize}
\item{For those values of $x$, for which the absolute value term is non-negative the absolute value can be omitted or
replaced by simple brackets, respectively.}
\item{For those values of $x$, for which the absolute value term is negative the term is bracketed and negated.}
\end{itemize}
\ \\
Then, the solution sets arising from the case analysis will be restricted as described in the 
\MSRef{VBKM02_FallBetrag}{previous module} and merged to the solution set of the initial inequality. 
\end{MInfo}

\begin{MExample}
To solve the absolute value inequality $|4x-2|<1$ two cases are distinguished:
\begin{itemize}
\item{For $x\geq \frac12$, the absolute value term is non-negative: 
In this case the inequality is equivalent to $(4x-2)<1$ or $x<\frac34$, respectively. 
Because of the case condition the solution set is only 
$\ML_1=\left[\frac12\MIntvlSep \frac34\MoIr[\right]$ in this case.}
\item{For $x<\frac12$, the absolute value term is negative: 
In this case the inequality is equivalent to $-(4x-2)<1$ or $x>\frac14$, respectively. 
Only the subset $\ML_2=\MoIl[\left] \frac14\MIntvlSep \frac12\MoIr[\right]$ 
satisfies the case condition and is the solution set.}
\end{itemize}
The union of the two solution intervals results in the solution set
$\ML=\MoIl[\left] \frac14\MIntvlSep \frac34\MoIr[\right]$ for the initial absolute value inequality:

\begin{center}
\MTikzAuto{%
\begin{tikzpicture}
% reelle Achse
\draw[->,color=black] (-1,0.0) -- (5,0.0);
\foreach \x in {-1, 0, 1, 2, 3, 4}
\draw[shift={(\x,0)},color=black] (0pt,2pt) -- (0pt,-2pt) node[below] {\footnotesize $\x$};
\draw (4.9,-0.3) node[] {$\mathbb{R}$};
% Enden:
\draw [line width=2.0pt,color=blue] (0.25,0.0)-- (0.75,0.0);
\draw [fill = white] (0.25,0) circle (1.5pt);
\draw [fill = white] (0.75,0) circle (1.5pt);
\end{tikzpicture}
}
\end{center}
\end{MExample}

\begin{MExercise}
To solve the absolute value inequality $|x-1|<2|x-1|+x$ two cases are distinguished:
\begin{MExerciseItems}
\item{On the interval \MLIntervalQuestion{20}{[1,infty)}{3}{UGL1}, both
terms in the absolute value terms are non-negative. 
The solution set of the inequality is in this case 
\MEquationItem{$\ML_1$}{\MLIntervalQuestion{20}{[1,infty)}{3}{UGL2}}.}
\item{On the interval \MLIntervalQuestion{20}{(-infty,1)}{3}{UGL3}, both
terms in the absolute value terms are negative. 
The solution set of the inequality is in this case
\MEquationItem{$\ML_2$}{\MLIntervalQuestion{20}{(-infty,1)}{3}{UGL4}}.}
\end{MExerciseItems}
The union of the two intervals results in the solution interval 
\MEquationItem{$\ML$}{\MLIntervalQuestion{25}{(-infty,infty)}{4}{UGL5}}.
\ \\ \ \\
\begin{MHint}{Solution}
For $x\in [1\MIntvlSep \infty\MoIr$, both terms in the absolute value terms are non-negative, 
one obtains the inequality $x-1<2(x-1)+x$, which is equivalent to $x>\frac12$. 
Because of the case condition one obtains $\ML_1=[1\MIntvlSep \infty\MoIr$ as solution set.
For $x\in\MoIl[\left] -\infty\MIntvlSep 1\MoIr[\right]$, 
both terms in the absolute value terms are negative and
one obtains $-(x-1)<-2(x-1)+x$. 
This inequality is equivalent to the inequality $x-1<x$ which is
always true. Thus,
the solution set for the second case is
$\ML_2=\MoIl[\left] -\infty\MIntvlSep 1\MoIr[\right]$.
\ \\ \ \\
Since $\ML=\ML_1\cup \ML_2=\R=\MoIl[\left] -\infty\MIntvlSep \infty\MoIr[\right]$ 
the inequality is always satisfied.
\end{MHint}
\end{MExercise}

\end{MIntro}

\begin{MXContent}{Quadratic Absolute Value Inequalities}{Quadratic Inequalities}{STD}
\MDeclareSiteUXID{VBKM03_QuadratischeUngleichungen}
\begin{MInfo}
An inequality is called \MEntry{quadratic}{inequality (quadratic)} in $x$ 
if it can be transformed into $x^2 + p x + q < 0$ (other comparing symbols are allowed).
\end{MInfo}
\ \\ \ \\
Hence, quadratic inequalities can be solved in two ways: by investigating the roots 
and the opening behaviour of the polynomial and by completing the square. Often completing
the square is simpler:
 

\begin{MInfo}
To solve an inequality by \MEntry{completing the square}{completing the square (inequalities)} 
one tries to transform it into the form $(x+a)^2<b$. Taking the square root then results
in the absolute value inequality $|x+a|<\sqrt{b}$ with the solution set 
$\MoIl[\left] -a-\sqrt{b}\MIntvlSep -a+\sqrt{b}\MoIr[\right]$ if $b\geq 0$. Otherwise 
the inequality is unsolvable.

The inverted inequality $|x+a|>\sqrt{b}$ has the solution set
$\MoIl[\left] -\infty\MIntvlSep -a-\sqrt{b}\MoIr[\right]\cup \MoIl[\left] -a+\sqrt{b}\MIntvlSep \infty\MoIr[\right]$. 
For $\leq$ and $\geq$ the corresponding boundary points have to be included.
\end{MInfo}

Always note the calculation rule $\sqrt{x^2}=|x|$ described in Modul~\MNRef{VBKM01}.

\begin{MExample}
Find the solution of the inequality $2x^2\geq 4x+2$. Collecting the terms on the left-hand side and dividing 
the inequality by $2$ results in $x^2-2x-1\geq0$. Completing the square on the 
left-hand side to the second binomial formula results in the equivalent inequality $x^2-2x+1\geq 2$
or $(x-1)^2\geq 2$, respectively. Taking the square root results in the absolute value 
inequality $|x-1|\geq\sqrt{2}$ with the solution set 
$\ML=\MoIl[\left] -\infty\MIntvlSep 1-\sqrt{2}\right]\cup \left[1+\sqrt{2}\MIntvlSep \infty\MoIr[\right]$.
\end{MExample}

On the other hand, the inequality $x^2-2x-1\geq0$ can be investigated as follows:
The left-hand side describes a parabola opened upwards. The roots $x_{1,2}=1\pm \sqrt2$ 
can be found using the $pq$ formula:

\begin{center}
%%\MUGraphicsSolo{parabelu.png}{width=0.4\linewidth}{width:400px}
\MTikzAuto{%
\begin{tikzpicture}[x=1.0cm, y=1.0cm,scale=1.50] 
\draw[black] (-1,0) -- (3,0) (0,-2) -- (0,2);
\foreach \x in {-1, 1, 2, 3}
\draw[shift={(\x,0)},color=black] (0pt,0pt) -- (0pt,-2.0pt) node[below=1.0pt] {\scriptsize $\x$};
\foreach \x in {-0.5, 0.5, ..., 3.0}
\draw[shift={(\x,0)},color=black] (0pt,0pt) -- (0pt,-1.0pt);
\foreach \y in {-2, -1, 1, 2}
\draw[shift={(0,\y)},color=black] (0pt,0pt) -- (-2.0pt,0pt) node[left=1.0pt] {\scriptsize $\y$};
\foreach \y in {-1.5, -0.5, 0.5, 1.5}
\draw[shift={(0,\y)},color=black] (0pt,0pt) -- (-1.0pt,0pt);
\draw[black] (-0.0pt,-0.0pt) node[anchor=north east] {\scriptsize $0$};
\clip(-1.0,-3.0) rectangle (3.0,2.0);
\draw[smooth,samples=21,domain=-1:3, line width=1.0pt,color=red!50!black] plot(\x,{\x*\x-2*\x-1});
\end{tikzpicture}
}
\end{center}
Since the parabola opens upwards, the inequality $x^2-2x-1\geq0$ is satisfied by the
values of $x$ in the parabola branches left and right to the roots, i.e. by the set 
$\ML=\MoIl[\left] -\infty\MIntvlSep 1-\sqrt{2}\right]\cup \left[1+\sqrt{2}\MIntvlSep \infty\MoIr[\right]$.

\begin{MInfo}
\MLabel{M03_InfoFormen}
Depending on the roots of $x^2+ p x + q$, the opening of the parabola and the 
comparing symbol, the quadratic inequality $x^2 +p x +q <0$ (including other comparing symbols) 
has one of the following solution sets:

\begin{itemize}
\item{the set of real numbers $\R$,}
\item{two branches $\MoIl[\left] -\infty\MIntvlSep x_1\MoIr[\right]\cup \MoIl[\left] x_2\MIntvlSep \infty\MoIr[\right]$ (including the boundary points for $\leq$ and $\geq$),}
\item{an interval $\MoIl x_1\MIntvlSep x_2\MoIr$ (including the boundary points for $\leq$ and $\geq$ if applicable),}
\item{a single point $x_1$,}
\item{the pointed set $\R\MSetminus\lbrace x_1\rbrace$,}
\item{the empty set $\lbrace\rbrace$.}
\end{itemize}
\end{MInfo}

Fill in the blanks in the following text describing the solution of a quadratic
inequality by investigating the behaviour of the parabola:

\begin{MExercise}
Find the solution set of the inequality $x^2+6x< -5$. 
Transformation results in the inequality \MLSimplifyQuestion{15}{x^2+6*x+5}{5}{x}{5}{1}{OBXP1}$<0$.
Using the $p q$ formula one obtains the set of roots
\MLParsedQuestion{9}{-1,-5}{3}{PXL}. The left-hand side
describes a parabola opening \MLQuestion{10}{upwards}{ObenX}.
It belongs to an inequality involving the comparing symbol $<$, hence 
the solution set is \MEquationItem{$\ML$}{\MLIntervalQuestion{15}{(-5,-1)}{5}{INVX}}.
\ \\ \ \\
\begin{MHint}{Solution}
Transformation results in $x^2+6x+5<0$. Using the $p q$ formula
one obtains the roots $x_{1,2}=-3\pm\sqrt{9-5}$, i.e.\ $x_1=-1$ and $x_2=-5$.
The left-hand side describes a parabola opening upwards. It satisfies the inequality 
involving $<$ only on the interval $\MoIl[\left] -5\MIntvlSep -1\MoIr[\right]$ excluding
the boundary points.
\end{MHint}
\end{MExercise}

\end{MXContent}

\begin{MXContent}{Further Types of Inequalities}{Further Types of Inequalities}{STD}
\MDeclareSiteUXID{VBKM03_WeitereUngleichungstypen}
Many other types of inequalities can be transformed into quadratic inequalities. Sometimes, 
case analyses have to be done or excluded values in the domain have to be observed:

\begin{MInfo}
An inequality containing \MEntry{fractions}{inequality (fractions)}, where the 
variable $x$ occurs in the denominator of composite terms, can be transformed into a 
form without fractions by multiplying the inequality by the least common denominator. 
However, in doing so, the roots of the denominators have to be excluded from the domain
of the new inequality. 

Additionally, if the inequality is multiplied by a term, different cases have to be distinguished 
depending on the sign of the term.
\end{MInfo}

\begin{MExample}
The inequality $2-\frac1x\leq x$ can be transformed by multiplying the inequality by $x$. Here, three 
cases have to be distinguished:
\begin{itemize}
\item{For $x>0$, the comparing symbol in the inequality is unchanged. The new inequality
reads $2x-1\leq x^2$ and is equivalent to $x^2-2x+1\geq 0$ or $(x-1)^2\geq 0$, respectively.
This inequality is always satisfied. Because of the case condition one obtains 
the solution set $\ML_1=\MoIl 0\MIntvlSep \infty\MoIr$.}
\item{For $x<0$, the comparing symbol in the inequality is inverted. The new inequality
reads $2x-1\geq x^2$ and is equivalent to $x^2-2x+1\leq 0$ or $(x-1)^2\leq 0$, respectively.
This inequality is only satisfied for $x=1$. But this value is excluded by the case condition, 
i.e.\ $\ML_2=\{\}$.}
\item{The single value $x=0$ is not in the domain of the initial inequality and hence it is
no solution.}
\end{itemize}

So, altogether one obtains the union set 
$\ML=\MoIl 0\MIntvlSep \infty\MoIr$ as solution set of the initial inequality.
\end{MExample}

Inequalities involving composite fraction and root terms often do not have solution
sets of the types described in info box~\MRef{M03_InfoFormen}:

\begin{MExample}
Find the solution set of the inequality $\sqrt{x}+\frac1{\sqrt{x}}>2$. 
The domain of the inequality is $\MoIl 0\MIntvlSep \infty\MoIr$.
Multiplying by $\sqrt{x}$ results in the inequality $x+1>2\sqrt x$. 
Here, no case analysis is required since $\sqrt{x}>0$ is in the domain.
Transformation results in $x-2\sqrt{x}+1>0$ or $(\sqrt{x}-1)^2>0$, respectively, 
which is satisfied for all $x\not=1$ in the domain.
Hence, the solution set of the initial inequality is
$\ML=\MoIl 0\MIntvlSep \infty\MoIr\MSetminus\lbrace 1\rbrace$:
\ \\ \ \\
\begin{center}
\MTikzAuto{%
\begin{tikzpicture}
% reelle Achse
\draw[->,color=black] (-1,0.0) -- (5,0.0);
\foreach \x in {-1, 0, 1, 2, 3, 4}
\draw[shift={(\x,0)},color=black] (0pt,2pt) -- (0pt,-2pt) node[below] {\footnotesize $\x$};
\draw (4.9,-0.3) node[] {$\mathbb{R}$};
% Enden:
\draw [line width=2.0pt,color=blue] (0,0.0)-- (1,0.0);
\draw [line width=2.0pt,color=blue] (1,0.0)-- (5,0.0);
\draw [fill = white] (0,0) circle (1.5pt);
\draw [fill = white] (1,0) circle (1.5pt);
\end{tikzpicture}
}
\end{center}

\end{MExample}

\end{MXContent}


\MSubsection{Final Test}
\MLabel{M03_Abschlusstest}

\begin{MTest}{Final Test Modul 3}
\MDeclareSiteUXID{VBKM03_Abschlusstest}

\begin{MExercise}
Find the value of the parameter $\alpha$ such that the inequality $2x^2\leq x-\alpha$ 
has exactly one solution:
\begin{MExerciseItems}
\item{The parameter value is \MEquationItem{$\alpha$}{\MLParsedQuestion{10}{1/8}{3}{PMA1}}.}
\item{In this case \MEquationItem{$x$}{\MLParsedQuestion{10}{1/4}{3}{PMA2}} is the only solution
of the inequality.}
\end{MExerciseItems}
\end{MExercise}


\begin{MExercise}
Find an absolute value function $g(x)$ describing the following graph as easy as possible.

%%\MUGraphics{abs3.png}{width=0.5\linewidth}{Funktionsgraph von $g(x)$.}{width:300px}
\begin{center}
\MTikzAuto{%
\begin{tikzpicture}[x=1.4cm, y=1.9cm] 
\draw[black] (-3,0) -- (3,0) (0,-3) -- (0,3);
\foreach \x in {-3, -2, -1, 1, 2, 3}
\draw[shift={(\x,0)},color=black] (0pt,0pt) -- (0pt,-3.0pt) node[below=1.0pt] {\normalsize $\x$};
\foreach \x in {-3.0, -2.8, ..., 3.0}
\draw[shift={(\x,0)},color=black] (0pt,0pt) -- (0pt,-1.5pt);
\foreach \y in {-3, -2, -1, 1, 2, 3}
\draw[shift={(0,\y)},color=black] (0pt,0pt) -- (-3.0pt,0pt) node[left=1.0pt] {\normalsize $\y$};
\foreach \y in {-3.0, -2.8, ..., 3.0}
\draw[shift={(0,\y)},color=black] (0pt,0pt) -- (-1.5pt,0pt);
%%\draw[black] (-0.0pt,-0.5pt) node[anchor=north east] {\small $0$};
\clip(-3.0,-3.0) rectangle (3.0,3.0);
\draw[black, line width=1.0pt,color=black] (-3,-3) -- (1,1) -- (2,4);
\end{tikzpicture}
}
\par
Graph of the function $g(x)$.
\end{center}
Try to find a representation of the form $g(x)=|x+a|+b x+c$. 
The kink in the graph indicates how the absolute value term looks like.

\begin{MExerciseItems}
\item{Find the solution set of the inequality $g(x)\leq x$ by means of the graph.\\
The solution set is \MEquationItem{$\ML$}{\MLIntervalQuestion{20}{(-infty;1]}{5}{AUX1}}.}
\item{\MEquationItem{$g(x)$}{\MLSimplifyQuestion{20}{abs(x-1)+2*x-1}{10}{x}{10}{0}{SIMPLE3}}. 
\\\MInputHint{Absolute values can be entered in the form \texttt{betrag(x-a)} or \texttt{abs(x-a)}.}}
\end{MExerciseItems}
\end{MExercise}

\begin{MExercise}
Which positive real numbers $x$ satisfy the following inequalities?
\begin{MExerciseItems}
\item{$|3x-6|\leq x+2$ has the solution set
\MEquationItem{$\ML$}{\MLIntervalQuestion{16}{[1,4]}{4}{COSH1}} (written as an interval).}
\item{$\frac{x+1}{x-1}\geq 2$ has the solution set 
\MEquationItem{$\ML$}{\MLIntervalQuestion{16}{(1,3]}{4}{COSH2}} (written as an interval).}
\end{MExerciseItems}
\MInputHint{Enter open intervals in the form $(3;5)$, closed intervals in the form 
$[3;5]$. Infinity can be entered a a word or shortly a \texttt{infty}. Do not use 
notation $]a;b[$ for open intervals. Sets can be entered by listing the elements
 $\lbrace 1;2;3\rbrace$. For the set brackets enter AltGr+7 or AltGr+0, respectively.}
\end{MExercise}

\end{MTest}

\newpage
\MPrintIndex

\end{document}
