\input{mintmod.tex}
\MPragma{MathSkip}

%\Mtikzexternalize

\begin{document}

\MSection{Equations in one Variable}
\MLabel{VBKM02}
\MSetSectionID{VBKM02} % wird fuer tikz-Dateien verwendet

\begin{MSectionStart}
\MDeclareSiteUXID{VBKM02_START}

Equations arise by equating two terms in which variables occur. Simple equations can be solved by 
applying transformations and solution formulas. For more sophisticated equations case analyses are 
required. This module consists of

\begin{itemize}
\item{Section \MNRef{M02_EinfacheGleichungen}: \MSRef{M02_EinfacheGleichungen}{Simple Equations},}
%\item{dem Abschnitt \MNRef{M02_Wurzelgleichungen}, \MSRef{M02_Wurzelgleichungen}{Gleichungen mit Wurzeln},}
\item{Section \MNRef{M02_Betragsgleichungen}: \MSRef{M02_Betragsgleichungen}{Absolute Value Equations},}
\item{and Section \MNRef{M02_Abschlusstest}: \MSRef{M02_Abschlusstest}{Final Test}.}
\end{itemize}
\end{MSectionStart}

\MSubsection{Simple Equations}
\MLabel{M02_EinfacheGleichungen}

\begin{MIntro}
\MDeclareSiteUXID{VBKM02_EinfacheGleichungenIntro}
\begin{MInfo}
An \MEntry{equation}{equation} is an expression of the form
$$
\text{left-hand side} \;=\; \text{right-hand side} \MDFPSpace
$$
with mathematical expressions on both sides of the equation. These expressions generally contain variables or unknowns (e.q.\ $x$). 
Depending on the variable values an equation is satisfied if both sides of the equation evaluate to the same value. An equation is not 
satisfied if the sides of the equation evaluate to different values.
\end{MInfo}

Equations describe relations between expressions or model a problem to be solved. An equation itself cannot
be true or false but some variables satisfy the equation and others do not. To test whether the equation is 
true or false for a single variable value this value has to be inserted into the equation. Then, both sides of the 
equation are evaluated to certain values. The equation is satisfied by an inserted variable value if the evaluated 
values coincide:

\begin{MExample}
The equation $2x-1=x^2$ has the right-hand side $x^2$ and the left-hand side $2x-1$. Inserting $x=1$ 
results in the value $1$ on both sides of the equals sign, hence $x=1$ is a solution of this equation. 
However, $x=2$ is no solution since the left-hand side of the equation is evaluated to the value $4$ while the 
right-hand side is evaluated to the value $3$.
\end{MExample}

\begin{MInfo}
The \MEntry{solution set}{solution set} $\ML$ of an equation is the set of all numbers satisfying the relation 
$$
\text{left-hand side} \;=\; \text{right-hand side} \MDFPSpace
$$
if inserted into the the equation instead of the variable (e.q. $x$).
\end{MInfo}

Typical problems concerning equations are:
\begin{itemize}
 \item{specify the solution set of an equation, i.e.\ find all variable values satisfying the equation,}
 \item{transform the equation, in particular, solve an equation for the variables, and}
 \item{find an equation modelling a problem described textually.}
\end{itemize}


\begin{MExample}
  A savings deposit is to be designed such that it gives a fixed return per year. The bank intends to achieve
  that the saver returns for a five years deposit exactly 600~Euro more than for a deposit of only two years. 

  First, the textual problem is translated into an equation with the variable $r$ denoting the return per year. Then, 
  the equation is $5\cdot r=2\cdot r+600$. It says that five payments (left-hand side of the equation) equal 
  two payments plus 600 (the unit Euro is then omitted during calculation).

  This equation can be solved for $r$ very easily by subtracting the term $2r$ to both sides of the equation. Then, the 
  equation reads $3r=600$ and dividing by $3$ results in the solution $r=200$.

  Thus, the bank has to offer a return of 200~Euro per year to reach the required savings target. 
\end{MExample}

% \begin{MExercise}
% Übersetzen Sie die folgenden Aussagen in eine Gleichung (ohne sie zu lösen):
% \begin{MExerciseItems}
% \item{In elf Jahren ist Max doppelt so alt wie jetzt ($t$ = jetziges Alter in Jahren):  ???}
% \item{Ein Kunde hat eine Rechnung über 1000 Euro zu bezahlen. Die monatlich zu zahlende Rate bleibt für ihn gleich unabhängig davon ob er
% \begin{itemize}
%  \item{die Rechnung mit $m$ gleichgroßen Raten abbezahlt,}
%  \item{vier Raten mehr als $m$ einteilt und zusätzlich 10 Euro Verzugsgebühren pro Monat an die Bank zahlt (die nicht zur Rechnungsbegleichung dienen).}
% \end{itemize}
% Ist $m$ die Anzahl der Monate, so gilt ???
% }
% \end{MExerciseItems}
% \end{MExercise}


\begin{MInfo}
Two equations are said to be \MEntry{equivalent}{equivalence} if they have the same solution set.

An \MEntry{equivalent transformation}{equivalent transformation} is a special transformation that changes
the equation but not its solution set. Important equivalent transformations are

\begin{itemize}
 \item{adding/subtracting terms to both sides of the equation,}
 \item{exchanging both sides of the equation,}
 \item{transformation of terms on one side of the equation, and}
 \item{inserting known relations.}
 \end{itemize}

The following transformations are considered as equivalent transformations only if the used term is non-zero 
(which can depend on the variables):

\begin{itemize}
 \item{multiplying/dividing by a term (this term has to be non-zero),}
 \item{taking the reciprocal on both sides of the equation (both sides have to be non-zero).}
\end{itemize}
\end{MInfo}

Here, the following \MEntry{notation}{equivalent transformation (notation)} is used:

\begin{itemize}
 \item{equivalent equations are indicated by the symbol $\Leftrightarrow$ (which reads: if and only if, i.e. one
  equation is satisfied if and only if the other equation is satisfied).}

\item{the symbol is underset by the transforming operation (or, for solutions with more than one line, the 
  transforming operation is written next to the transformation).}
\end{itemize}

Importantly, the reader should be able to understand which transformation was carried out.

\begin{MExample}
This example illustrates two simple single-lined equivalent transformations. Even though the symbol $\Leftrightarrow$ 
is two-sided the notation is interpreted in such a way that the transformation is applied from 
left to right:
$$
3x-x^2 \;=\; 2x-x^2+1\;\;\MUnderset{+x^2}\Leftrightarrow\;\;  3x\;=\; 2x+1 \;\;\MUnderset{-2x}\Leftrightarrow\;\; x \;=\; 1 \MDFPeriod
$$
The left equation and the right equation are equivalent. On the left we have the initial equation
(corresponding to a certain textual problem) and on the right we have an equivalent equation 
showing the solution immediately.
\end{MExample}

\begin{MExample}
\MLabel{BSP_Umformungen1}
For several complicated transformations the transformation steps should be written below each other.
In this case the respective transformations are separated from the equation by vertical bars. 

\begin{eqnarray*}
& \text{Start:} & 12+t \;=\;\Mdfrac{2t}{2t^2}+t \ \ \ \ \MSep \ -t\ \\ \ \\
& \Leftrightarrow & 12 \;=\; \Mdfrac{2t}{2t^2}  \ \ \ \ \MSep \ \text{sides exchanged}\ \\ \ \\
& \Leftrightarrow & \Mdfrac{2t}{2t^2} \;=\; 12  \ \ \ \ \MSep \ \text{left-hand side transformed}\ \\ \ \\
& \Leftrightarrow & \Mdfrac1t \;=\; 12  \ \ \ \ \MSep \ \text{reciprocals taken}\ \\ \ \\
& \Leftrightarrow & t \;=\; \Mdfrac{1}{12} \MDFPeriod
\end{eqnarray*}

Here, after the vertical bar both short symbols as, e.g. $-t$, and textual descriptions are allowed. 
Importantly, the reader should be able to understand which transformations were carried out and should be able to decide whether 
they are correct.

\end{MExample}
\end{MIntro}

\begin{MXContent}{Conditions in Transformations}{Conditions}{STD}
\MDeclareSiteUXID{VBKM02_Bedingungen}
Multiplication, division, and taking reciprocals are only equivalent transformations if the factors or terms are 
non-zero. In example~\MRef{BSP_Umformungen1}, the reader understands that both sides of the equation are non-zero 
such that the transformation is allowed. If the variables themselves are used in the transformation it has to be 
noted that the respective term has to be non-zero. The solution at the end of the transformations is then 
only valid for variable values satisfying the transformation conditions. All other values have to be checked 
\textit{separately}, typically by inserting the value into the equation:

\begin{MExample}
In this example, the necessary transformation conditions are not problematic:
\begin{eqnarray*}
& \text{Start:} & 9x \;=\;81x^2  \ \ \ \ \MSep \ :x\text{, transformation allowed if }x\not=0\ \\ \ \\
& \Leftrightarrow & 9 \;=\; 81x  \ \ \ \ \MSep \ :81\text{ and exchange sides}\ \\ \ \\
& \Leftrightarrow & x \;=\; \Mdfrac19 \ \ \ \ \text{and this value satisfies the condition }x\not=0 \MDFPeriod
\end{eqnarray*}
The value $x=0$, initially rejected by the transformation condition, has to be checked separately. The equation $9x=81x^2$ is 
also satisfied for $x=0$, hence $x=0$ is also a solution of the equation. In set notation, the 
solution set is $\ML=\lbrace 0\MElSetSep \Mtfrac19\rbrace$.
\end{MExample}

Anyway, values violating a condition have to be checked separately, in particular, they can be finally 
part of the solution.

\begin{MExample}
\begin{eqnarray*}
& \text{Start:} & x^2-2x \;=\; 2x-4 \ \ \ \ \MSep \ \text{factor out on both sides} \ \\ \ \\
& \Leftrightarrow & x\cdot (x-2) \;=\; 2\cdot (x-2) \ \ \ \ | \ :(x-2)\text{, transformation only allowed if }x\not=2\ \\ \ \\
& \Leftrightarrow & x\;=\; 2  \MDFPeriod
\end{eqnarray*}
This value of $x$ violates the condition $x\not=2$. Hence, this is possibly no solution. Inserting $x=2$
into the initial equation gives $2^2-2\cdot 2=0$ on the left-hand side and also $2\cdot 2-4=0$ on the 
right-hand side. Hence, $x=2$ is indeed a solution, even though it violated the transformation condition.
\end{MExample}

\begin{MExercise}
Find the solution of the equation $(x-2)(x-3)=x^2-9$ by transforming the right-hand side using the third 
binomial formula and then dividing by a common factor. 

The solution is $x$ = \MLParsedQuestion{5}{3}{5}{EASY1}.

\begin{MHint}{Solution}
The correct transformation steps including conditions are
\begin{eqnarray*}
& \text{Start:} & (x-2)(x-3)\;=\; x^2-9 \ \ \ \ \MSep \ \text{transformation of the right-hand side} \ \\ \ \\
& \Leftrightarrow & (x-2)(x-3) \;=\; (x+3)(x-3) \ \ \ \ \MSep \ :(x-3)\text{, transformation allowed if }x\not=3\ \\ \ \\
& \Leftrightarrow & x-2\;=\; x+3 \ \ \ \ \MSep \ -x \ \\ \ \\
& \Leftrightarrow & -2 \;=\; 3 \; \text{is a wrong equation.}
\end{eqnarray*}
Importantly, this equation is only wrong for $x\not=3$. We have to check $x=3$ separately, and indeed 
$x=3$ satisfies the initial equation.
\end{MHint}
\end{MExercise}
\end{MXContent}

\begin{MXContent}{Proportionality and Rule of Three}{Proportionality}{STD}
\MLabel{VBKM02_Dreisatz}
\MDeclareSiteUXID{VBKM02_Dreisatz}
In practise, a relation between two quantities that occurs frequently is the 
\MEntry{proportionality}{proportionality} between two quantities, e.g. between 
mass and volume, time and travelled distance or weight (quantity) of a product and its price. 
The relation can be exemplary for certain fixed quantities. Then, a first aim is to formulate the 
resulting relation for another application example. 
The procedure shall be illustrated by an example. 


\begin{MExample}
$5\MEinheit{kg}$ of apples cost $3$ Euro. How much do $11\MEinheit{kg}$
of apples cost?
\par
The initial relation can be formulated as follows:
$$
5\MEinheit{kg} \MDFPSpace \MRelates \MDFPSpace 3\MBlank \text{Euro}
\MDFPeriod
$$
It is assumed that these quantities are proportional to each other. In the next
step the relation between the quantities is reduced to a unit of one of the quantities, 
namely to the unit of the given quantity. Here, both quantities 
are multiplied by $1/5$ -- i.e. the unit is $1\MEinheit{kg}$ --:
$$
1\MEinheit{kg} \MDFPSpace \MRelates \MDFPSpace 
\frac{1}{5}\cdot 3\MBlank \text{Euro} = \MZahl{0}{6}\MBlank \text{Euro}
\MDFPeriod
$$
Finally, both sides of the equation are multiplied by the multiple of the 
respective unit of the specified quantity, in this case by the factor $11$: 
$$
11\MEinheit{kg} \MDFPSpace \MRelates \MDFPSpace 
11\cdot\MZahl{0}{6}\MBlank \text{Euro} = \MZahl{6}{6}\MBlank \text{Euro}
\MDFPeriod
$$
The required price for $11\MEinheit{kg}$ of apples is therefore
$\MZahl{6}{60}\MBlank \text{Euro}$.
\end{MExample}

We have derived the required relation by deriving a relation for one unit of a quantity from 
the initial relation. This procedure demonstrated here as an example is called 
\MEntry{rule of three}{rule of three}.

The posed problem can also be solved by introducing a proportionality factor. 
Again, we consider the example above.

\begin{MExample}
The price $P$ is proportional to the mass $m$. Hence, it exists a constant 
$k$ with
$$
P=k m \MDFPeriod
$$
Since this relation also holds for the given values $m_0=5\MEinheit{kg}$
and $P_0=3\MBlank\text{Euro}$ it follows 
\begin{eqnarray*}
  P_0=k m_0 & \MTSP\MSep\MTSP & \text{multiplying by}\MBlank \frac{1}{m_0} \\
  \Longleftrightarrow\MDFPSpace\frac{P_0}{m_0}=k &;& 
\end{eqnarray*}
hence in this case
$$
k=\frac{3}{5} = \MZahl{0}{6} \MDFPSpace,
$$
taken in the unit of Euro per kg. (As a scientist you would correctly write 
 $k=\MZahl{0}{6} \MBlank \text{Euro}/\MEinheit[]{kg}$, since proportionality factors 
generally carry a dimensional unit.) Using $m_1=11\MEinheit{kg}$, one obtains 
finally
$$
P_1=k m_1 = \MZahl{0}{6}\cdot 11 =\MZahl{6}{6} \MBlank \text{(Euro)}
\MDFPSpace
$$
which is the same result as for using the rule of three (see previous example).
\end{MExample}

\begin{MExercise}
A car takes $9$~minutes to travel a distance of $6\MEinheit{km}$.
\begin{MExerciseItems}
\item{%
Which distance $s$ the car travels within $15$~minutes?
\medskip\par
The solution is $s_{15}$ = \MLParsedQuestion{5}{10}{5}{EASY2}$\MEinheit{km}$.
}
\item{%
The proportionality factor between travelled distance $s$ and travelling 
time $t$ is the velocity $v$ of the car.
\medskip\par
The velocity is $v$ = \MLParsedQuestion{5}{40}{5}{EASY3}$\MEinheit{km}/\MEinheit{h}$.
}
\end{MExerciseItems}
\begin{MHint}{Solution}
>From the given values we know that the car travels $\frac{6}{9}\MEinheit{km}
=\frac{2}{3}\MEinheit{km}$  within one minute and therefore 
$15\cdot\frac{2}{3}\MEinheit{km}=10\MEinheit{km}$ within $15$~minutes.
\par
So, the velocity is
$$v=\frac{10\MEinheit{km}}{15\MEinheit{min}} = 
\frac{10\MEinheit{km}}{(1/4)\MEinheit{h}} =
40\MEinheit{}\frac{\MEinheit[]{km}}{\MEinheit[]{h}}
\MDFPeriod
$$
\end{MHint}
\end{MExercise}
\end{MXContent}

\begin{MXContent}{Solving linear Equations}{Solving}{STD}
\MLabel{VBKM02_LineareGleichungenLoesen}
\MDeclareSiteUXID{VBKM02_Aufloesen}
\begin{MInfo}
A \MEntry{linear equation}{equation (linear)} is an equation in which only multiples of 
variables and constants occur.

For a linear equation in one variable (here the variable $x$) one of the following 
three statements holds:
\begin{itemize}
 \item{The equation has no solution.}
 \item{The equation has a single solution.}
 \item{Every value of $x$ is a solution of the equation.}
\end{itemize}
\end{MInfo}

These three cases are distinguished by means of the transformation steps:
\begin{itemize}
\item{If the transformation ends up in a statement that is wrong for all $x$ (e.g. $1=0$) 
then the equation is unsolvable.}
\item{If the transformation ends up in a statement that is true for all $x$ (e.g. $1=1$)
then the equation is solvable for all values of $x$.}
\item{Otherwise, the equation can be solved, i.e. it can be transformed into 
the equation $x=\text{value}$ which is the solution.}
\end{itemize}

\begin{MXInfo}{Set notation}
Using the set notation (with the solution set $\ML$) these cases can be expressed as follows:
\begin{itemize}
 \item{$\ML=\lbrace\rbrace$ or $L=\MEmptyset$ if there is no solution,}
 \item{$\ML=\lbrace \text{value}\rbrace$ if there is a single solution,}
 \item{$\ML=\R$ if all real numbers $x$ are a solution.}
\end{itemize}
\end{MXInfo}


\begin{MExample}
The linear equation $3x+2=2x-1$ has one solution. 
This solution is obtained by equivalent transformations:
$$
3x+2 \;=\; 2x-1 \;\;\MUnderset{-2x}\Leftrightarrow\;\; x+2\;=\;-1\;\;\MUnderset{-2}\Leftrightarrow\;\; x\;=\; -3 \MDFPeriod
$$
Hence, $x=-3$ is the only solution.
\end{MExample}

\begin{MExample}
The linear equation $3x+3=9x+9$ has the solution:
$$
3x+3 \;=\; 9x+9 \;\;\MUnderset{:(x+1)}\Leftrightarrow\;\; 3\;=\;9 \MDFPeriod
$$
This statement is wrong. Hence, for all $x\not=-1$ (transformation condition) the equation is wrong.
Inserting $x=-1$ satisfies the equation, and so the only solution is $x=-1$.

Alternatively, the equation could have been transformed as follows:
$$
3x+3 \;=\; 9x+9 \;\;\MUnderset{-3x-9}\Leftrightarrow\;\; -6 \;=\; 6x \;\; \Leftrightarrow\;\; x \;=\; -1 \MDFPeriod
$$
\end{MExample}

\begin{MExercise}
Transform the following linear equations and specify their solution sets:
\MInputHint{Enter simply \texttt{$\lbrace a\rbrace$} for a unit set and $\lbrace\rbrace$ for an empty set.}\ \\
\begin{MExerciseItems}
\item{The equation $x-1=1-x$ has the solution set \MEquationItem{$\ML$}{\MLParsedQuestion{4}{1,1}{4}{LUA1}},}
\item{The equation $4x-2=2x+2$ has the solution set \MEquationItem{$\ML$}{\MLParsedQuestion{4}{2,2}{4}{LUA2}},}
\item{The equation $2x-6=2x-10$ has the solution set \MEquationItem{$\ML$}{\MLParsedQuestion{4}{}{4}{LUA3}}.}
\end{MExerciseItems}

\begin{MHint}{Solution}
The first equation can be transformed into $2x=2$ or $x=1$, respectively, so 
the solution set is $\ML=\lbrace 1\rbrace$. The second equation can be transformed into $2x=4$ and the solution set is 
$\ML=\lbrace 2\rbrace$. The third equation can be transformed into $-6=-10$ which is a false statement, 
hence $\ML=\lbrace\rbrace$.
\end{MHint}
\end{MExercise}

\begin{MExercise}
Find the solution of the general linear equation $a x=b$ with $a$ and $b$ being real numbers.
Specify the values of $a$ and $b$ for which the following three cases occur:
\begin{itemize}
 \item{Every value of $x$ is a solution ($\ML=\R$) if 
\MEquationItem{$a$}{\MLParsedQuestion{4}{0}{4}{ALG1}} and $b=0$.}
 \item{There is no solution ($\ML=\MEmptyset$) if \MEquationItem{$a$}{\MLParsedQuestion{4}{0}{4}{ALG2}} 
and $b\not=$\MLParsedQuestion{4}{0}{4}{ALG3}.}
 \item{Otherwise, there is a single solution, namely 
\MEquationItem{$x$}{\MLSimplifyQuestion{5}{b/a}{10}{a,b}{10}{513}{VBKM02ALTFALL3}}.}
\end{itemize}

\begin{MHint}{Solution}
Every value of $x$ is a solution ($\ML=\R$) if $a=0$ and $b=0$.
There is no solution ($\ML=\MEmptyset$) if $a=0$ and $b\not=0$.
Otherwise, there is only one solution, namely $x=\Mtfrac{b}{a}$.
\end{MHint}

\end{MExercise}


\end{MXContent}

\begin{MXContent}{Solving quadratic Equations}{Quadratic Equations}{STD}
\MLabel{VBKM02_QuadratischeGleichungen}
\MDeclareSiteUXID{VBKM02_Quadratische Gleichungen}
\begin{MInfo}
A \MEntry{quadratic equation}{equation (quadratic)} is an equation of the form
  $a x^2 + b x + c = 0$ with $a\not=0$, or, in reduced form, $x^2+ p x + q=0$. 
This form is obtained by dividing the equation by $a$.

For a quadratic equation in one variable (here the variable $x$) one of the following three statements 
holds:
\begin{itemize}
 \item{The quadratic equation has no solution: $\ML=\lbrace\rbrace$.}
 \item{The quadratic equation has a single solution $\ML=\lbrace x_1\rbrace$.}
 \item{The quadratic equation has two different solutions $\ML=\lbrace x_1\MElSetSep x_2\rbrace$.}
\end{itemize}
\end{MInfo}

The solutions are obtained by applying \MEntry{quadratic solution formulas}{solution formulas}.

\begin{MInfo}
\MLabel{VBKM02_pqFormel}
The \MEntry{$p q$ formula}{pq formula} for solving the equation $x^2+p x + q = 0$ reads
$$
x_{1,2} \;=\; -\Mdfrac{p}{2}\pm \sqrt{\Mdfrac14p^2-q} \MDFPeriod
$$
Here, the equation has
\begin{itemize}
\item{no (real) solution if $\Mtfrac14p^2-q<0$ (taking the square root is not allowed),}
\item{a single solution $x_1=-\Mtfrac{p}{2}$ if $\Mtfrac14p^2=q$ and the square root is zero,}
\item{two different solutions if the square root is a positive number.}
\end{itemize}

The expression $D:=\Mtfrac14p^2-q$ underneath the square root considered above is called 
\textbf{discriminant}.
\end{MInfo}
The solution of a quadratic equation is often described by an alternative formula:
\begin{MInfo}
\MLabel{VBKM02_abcFormel}
For the equation $a x^2+b x + c = 0$ with $a\ne 0$ the
\MEntry{$a b c$ formula}{abc formula} reads
$$
x_{1,2} \;=\; \frac{-b\pm\sqrt{b^2-4 a c}}{2a} \MDFPeriod
$$
Here, the equation has
\begin{itemize}
\item{no (real) solution if $b^2-4 a c<0$ (the square root of a negative number is undefined within the
range of real numbers),}
\item{a single solution $x_1=-\Mtfrac{b}{2a}$ if $b^2=4 a c$ and the square root is zero,}
\item{two different solutions if the square root is a positive number.}
\end{itemize}
Again, the expression $D:=b^2-4 a c$ underneath the square root considered above is called 
\textbf{discriminant}.
\end{MInfo}
Certainly, both formulas result in the same solutions. (For applying the $pq$ formula, the quadratic 
equation is to be divided by the factor $a$ of the quadratic term.)
\medskip\par
This three different cases correspond to three different numbers of intersection points between 
the graph of a parabola opening upwards $f(x)=x^2+p x+ q$ and the $x$ axis (applying the $p q$ formula).

\begin{center}
%%\MUGraphicsSolo{para1b.png}{width=0.2\linewidth}{width:230px}\ \ 
%%\MUGraphicsSolo{para2b.png}{width=0.2\linewidth}{width:230px}\ \ 
%%\MUGraphicsSolo{para3b.png}{width=0.2\linewidth}{width:230px}
\MTikzAuto{%
\begin{tikzpicture}[x=1.0cm, y=1.0cm,scale=0.60] 
\foreach \sx/\fsy in {-7.0cm/0.5,0.0cm/0.0,7.0cm/-0.5} {
\begin{scope}[xshift=\sx,yshift=0]
\draw[black] (-3,0) -- (3,0) (0,-3) -- (0,3);
\foreach \x in {-3, -2, -1, 1, 2, 3}
\draw[shift={(\x,0)},color=black] (0pt,0pt) -- (0pt,-2.0pt) node[below=1.0pt] {\tiny $\x$};
\foreach \x in {-2.5, -1.5, ..., 2.5}
\draw[shift={(\x,0)},color=black] (0pt,0pt) -- (0pt,-1.0pt);
\foreach \y in {-3, -2, -1, 1, 2, 3}
\draw[shift={(0,\y)},color=black] (0pt,0pt) -- (-2.0pt,0pt) node[left=1.0pt] {\tiny $\y$};
\foreach \y in {-2.5, -1.5, ..., 2.5}
\draw[shift={(0,\y)},color=black] (0pt,0pt) -- (-1.0pt,0pt);
\draw[black] (-0.0pt,-0.0pt) node[anchor=north east] {\tiny $0$};
\clip(-3.0,-3.0) rectangle (3.0,3.0);
\draw[smooth,samples=27,domain=-3:3, line width=1.0pt,color=red!50!black] plot(\x,{\x*\x+\fsy});
\end{scope}
}
\end{tikzpicture}
}
\end{center}

Three cases: no intersection point, one intersection point and two intersection points with the
$x$ axis.


\begin{MExample}
The quadratic equation $x^2-x+1=0$ has no solution since the discriminant $\Mtfrac14p^2-q=-\Mtfrac34$
within the $p q$ formula is negative.
In contrast, the equation $x^2-x-1=0$ has two solutions
\begin{eqnarray*}
x_1 &=& \Mdfrac12+\sqrt{\Mdfrac14+1} \;=\; \Mdfrac12(1+\sqrt5) \MDFPSpace ,\ \\
x_2 &=& \Mdfrac12-\sqrt{\Mdfrac14+1} \;=\; \Mdfrac12(1-\sqrt5) \MDFPeriod
\end{eqnarray*}
\end{MExample}

\begin{MInfo}
\MLabel{VBKM02_Scheitelpunktform}
%%Eine quadratische Gleichung ist in \MEntry{Scheitelpunktform}{Scheitelpunktform}, wenn sie die Form $a\cdot (x-s)^2=d$ besitzt mit $a\not=0$ und $d\geq 0$.
%%Für den Funktionsausdruck der zugehörigen Parabel liest sich diese Form $f(x)=a\cdot (x-s)^2-d$.
%%In dieser Situation ist $(s\,|\,-d)$ der \MEntry{Scheitelpunkt}{Scheitelpunkt (Parabel)} der Parabel.
%%
%%Falls $a>0$ ist gibt es zwei Lösungen
%%$$
%%x_1 \;=\; s-\sqrt{\Mdfrac{d}{a}} \MDFPSpace,\MDFPaSpace x_2 \;=\; s+\sqrt{\Mdfrac{d}{a}}
%%$$
%%der Gleichung, diese liegen symmetrisch um die $x$-Koordinate des Scheitelpunkts. Für $d=0$ gibt es nur eine Lösung.
%%
%%Das Vorzeichen von $a$ bestimmt, ob die Gleichung eine nach oben oder unten geöffnete Parabel beschreibt.
The function expression of a parabola has \MEntry{vertex form}{vertex form} if the function
has the form $f(x)=a\cdot (x-s)^2-d$ with $a\ne 0$. In this case, $\MPointTwo{s}{-d}$
is the \MEntry{vertex}{vertex (parabola)} of the parabola. The corresponding 
quadratic equation for $f(x)=0$ then reads $a\cdot (x-s)^2=d$.

Dividing this equation by $a$ one obtains the equivalent quadratic equation 
$(x-s)^2=\frac{d}{a}$. Since the left-hand side is a square of a real number, only 
solutions exist if and only if the right-hand side is non-negative as well, i.e.
$\frac{d}{a}\ge 0$. By taking the square root, taking the two possible signs into
account, one obtains $x-s=\pm\sqrt{\frac{d}{a}}$.

So, for $\frac{d}{a}>0$ two solutions of the equation exist:
$$
x_1 \;=\; s-\sqrt{\Mdfrac{d}{a}} \MDFPSpace,\MDFPaSpace x_2 \;=\; s+\sqrt{\Mdfrac{d}{a}}\,;
$$
they are symmetric to the $x$ coordinate $s$ of the vertex. For $d=0$, only one
solution exists.

The sign of $a$ determines whether the function expression describes a parabola 
opening upwards or downwards. 
\end{MInfo}

The quadratic equation has only one single solution $s$ if it can be transformed into the form
$(x-s)^2=0$.

\begin{MInfo}
\MLabel{VBKM02_QuadratischErgaenzung}

Any quadratic equation can be transformed (after collecting terms on the left-hand side
and normalisation, if necessary) to vertex form by 
\MEntry{completing the square}{completing the square}. For this, a constant is
added to both sides of the equation such that on the left-hand side we have a term of the 
form $x^2\pm 2s x+s^2$ to which the first or second binomial formula can be applied.
\end{MInfo}

\begin{MExample}
Adding the constant $2$ transforms the equation $x^2-4x+2=0$ into the
form $x^2-4x+4=2$ or into the form $(x-2)^2=2$, respectively. From this,
the two solutions $x_1=2-\sqrt{2}$ and $2+\sqrt{2}$ can be seen immediately.
In contrast, the quadratic equation $x^2+x=-2$ has no solution since completing
the square results in $x^2+x+\Mtfrac14=-\Mtfrac74$ or $(x+\Mtfrac12)^2=-\Mtfrac74$, respectively, 
where the right-hand side is negative for $a=1$.
\end{MExample}

\begin{MExercise}
Find the solutions of the following quadratic equations by completing the square after 
collecting terms on the left-hand side and normalisation (i.e.\ selecting $a=1$):

\begin{MExerciseItems}
\item{$x^2=8x-1$ has the vertex form \MEquationItem{\MLSimplifyQuestion{20}{x^2-8*x+16}{5}{x}{5}{0}{MSPF1}}{\MLParsedQuestion{5}{15}{5}{GLD1}}.\\The solution set is \MEquationItem{$\ML$}{\MLParsedQuestion{30}{4-sqrt(15),4+sqrt(15)}{5}{SQRT1}}.}
\item{$x^2=2x+2+2x^2$ has the vertex form \MEquationItem{\MLSimplifyQuestion{20}{x^2+2*x+1}{5}{x}{5}{0}{MSPF2}}{\MLParsedQuestion{5}{-1}{5}{GLD2}}.\\The solution set is \MEquationItem{$\ML$}{\MLParsedQuestion{30}{}{5}{SQRT2}.}}
\item{$x^2-6x+18=-x^2+6x$ has the vertex form \MEquationItem{\MLSimplifyQuestion{20}{x^2-6*x+9}{5}{x}{5}{0}{MSPF3}}{\MLParsedQuestion{5}{0}{5}{GLD3}}.\\The solution set is \MEquationItem{$\ML$}{\MLParsedQuestion{30}{3,3}{5}{SQRT3}.}}
\end{MExerciseItems}
\MInputHint{Enter sets in the form \texttt{$\lbrace$ a;b;c;$\ldots\rbrace$}. Enter the empty set as $\lbrace\rbrace$.}

\begin{MHint}{Solution}
The transformations are for the first equation
\begin{eqnarray*}
&&x^2=8x-1 \ \\
&\Leftrightarrow&x^2-8x+1=0 \ \\
&\Leftrightarrow&x^2-8x+16=15 \ \\
&\Leftrightarrow&(x-4)^2=15 \ \\
&& \ML=\lbrace 4-\sqrt{15}\MElSetSep 4+\sqrt{15}\rbrace
\end{eqnarray*}
and for the second equation
\begin{eqnarray*}
&&x^2=2x+2+2x^2 \ \\
&\Leftrightarrow&x^2+2x+2=0 \ \\
&\Leftrightarrow&x^2+2x+1=-1 \ \\
&\Leftrightarrow&(x+1)^2=-1\ \\
&& \ML=\lbrace\rbrace
\end{eqnarray*}
and for the third equation
\begin{eqnarray*}
&&x^2-6x+18=-x^2+6x \ \\
&\Leftrightarrow&2x^2-12x+18=0 \ \\
&\Leftrightarrow&x^2-6x+9=0 \ \\
&\Leftrightarrow&(x-3)^2=0\ \\
&& \ML=\lbrace3\rbrace \MDFPeriod
\end{eqnarray*}
\end{MHint}

\end{MExercise}

\end{MXContent}


\MSubsection{Absolute Value Equations}
\MLabel{M02_Betragsgleichungen}

\begin{MIntro}
\MDeclareSiteUXID{VBKM02_Betragsgleichungen_Intro}
\MLabel{VBKM02_Betrag}
The absolute value $|x|$ assigns a variable $x\in\R$ its value without sign: If $x\geq 0$, then $|x|=x$, 
otherwise $|x|=-x$ (see figure).

\begin{center}
%%\MUGraphicsSolo{betrag1.png}{width=0.4\linewidth}{width:350px}\\
\MTikzAuto{%
\begin{tikzpicture}[x=1.0cm, y=1.0cm] 
\draw[black] (-3,0) -- (3,0) (0,-3) -- (0,3);
\foreach \x in {-3, -2, -1, 1, 2, 3}
\draw[shift={(\x,0)},color=black] (0pt,0pt) -- (0pt,-2.0pt) node[below=1.0pt] {\scriptsize $\x$};
\foreach \x in {-2.5, -1.5, ..., 2.5}
\draw[shift={(\x,0)},color=black] (0pt,0pt) -- (0pt,-1.0pt);
\foreach \y in {-3, -2, -1, 1, 2, 3}
\draw[shift={(0,\y)},color=black] (0pt,0pt) -- (-2.0pt,0pt) node[left=1.0pt] {\scriptsize $\y$};
\foreach \y in {-2.5, -1.5, ..., 2.5}
\draw[shift={(0,\y)},color=black] (0pt,0pt) -- (-1.0pt,0pt);
\draw[black] (-0.0pt,-0.5pt) node[anchor=north east] {\scriptsize $0$};
\clip(-3.0,-3.0) rectangle (3.0,3.0);
\draw[black, line width=1.0pt,color=red!50!black] (-3,3) -- (0,0) -- (3,3);
\end{tikzpicture}
}
\par
The absolute value $|x|$ as a function of $x$.
\end{center}

Absolute value equations are equations in which one absolute value or several absolute values occur. 
Problems arise since the absolute value is calculated by distinguishing the two cases  
$$
\left|{\,\text{term}\,}\right| \;\; =\;\; \MCases{\text{term} & \text{for}\;\text{term}\geq 0\\ -\text{term} & \text{for}\;\text{Term}<0}\,.
$$
For solving absolute value equations, these cases have to be solved step by step 
and analysed to find the solutions.

\begin{MExample}
Obviously, the absolute value equation $|x|=2$ has the solution set $\ML=\lbrace 2\MElSetSep -2\rbrace$.
Just as easy, it can be seen that $|x-1|=3$ has the solution set $\ML=\lbrace -2\MElSetSep 4\rbrace$.
\end{MExample}

As soon as beside the absolute value several other terms occur a case analysis is required.
In the following section we will explain in detail how this analysis is done and how it is 
written correctly since the case analysis will play an important role in the next modules.
\end{MIntro}

\begin{MXContent}{Carry out a Case Analysis}{Case Analysis}{STD}
\MDeclareSiteUXID{VBKM02_Fallunterscheidungen}

\begin{MInfo}
\MLabel{VBKM02_FallBetrag}
To solve an \MEntry{absolute value equation}{absolute value equation} two cases are distinguished:
\begin{itemize}
\item{For all values of $x$ for which the absolute value term is non-negative the absolute value can be omitted or
replaced by simple brackets, respectively.}
\item{For all values of $x$ for which the absolute value term is negative the term is bracketed and negated.}
\end{itemize}
Then, the solution sets from the case analyses will be restricted to satisfy the case conditions. 
Only if this procedure is finished for all cases, the solution subsets will be merged to the 
solution set of the initial equation. 
\end{MInfo}

For solving absolute value equations it is important to write down the solution steps correctly
and to distinguish the cases clearly.

The following video demonstrates a detailed written solution of the absolute value
equation $|2x-4|=6$ by case analysis.


\MVideo{vidbsp1}{Carry out a case analysis.\MCopyrightLabel{VBKM06_Video_Beispiel1}}
\ \\
\MCopyrightNotice{\MCCLicense}{FSZ}{MINT}{Dozentin: Dipl.-ing. Heike Herold}{VBKM06_Video_Beispiel1}

The case analysis presented in the video reads shortly
$$
|2x-4| \;=\; \MCases{2x-4 &\text{for}\;x\geq 2 \\ -2x+4 & \text{for}\;x<2} \;=\;\MCases{2x-4 &\text{for}\;x\geq 2 \\ -2x+4 & \text{otherwise}} \MDFPeriod
$$
\MInputHint{In einem Eingabefeld würde man das als \texttt{falls(x>=2;2*x-4;-2*x+4)} eintippen. Auch \texttt{falls(x<2;-2*x+4;2*x-4)} wäre richtig.}

\begin{MExercise}
Describe the values of the expression $2\cdot |x-4|$ by a case analysis:
\begin{center}
$\displaystyle 2\cdot |x-4|$ = \MLSimplifyQuestion{30}{falls(x>=4,2*(x-4),-2*(x-4))}{15}{x}{4}{128}{VBKM02ALTFALL1}.
\end{center}
\MInputHint{Enter the case analysis in the form \texttt{falls(BEDINGUNG;W1;W2)}, where \texttt{W1} is the value of the expression if the corresponding condition is satisfied. Do not use the absolute value function.}
\begin{MHint}{Solution}
\begin{eqnarray*}
2\cdot |x-4| \;=\; \MCases{2x-8 &\text{for}\;x\geq 4 \\ -2x+8 & \text{for}\;x<4}
\end{eqnarray*}
\end{MHint}
\end{MExercise}

\begin{MExercise}
Reproduce the steps shown in the video \MRef{VBKM06_Video_Beispiel1} to solve the absolute 
value equation $|6+3x|=12$.

The case analysis reads shortly $|6+3x|$ = \MLSimplifyQuestion{40}{for(x>=-2,6+3*x,-6-3*x)}{15}{x}{3}{128}{VBKM02ALTFALL2}.
\begin{MHint}{Solution}
\begin{eqnarray*}
|6+3x| \;=\; \MCases{6+3x &\text{for}\;x\geq -2 \\ -6-3x & \text{for}\;x<-2}
\end{eqnarray*}
\end{MHint}\\
\MInputHint{Enter the case analysis in the form \texttt{falls(BEDINGUNG;W1;W2)}, where \texttt{W1} is the value of the expression if the corresponding condition is satisfied.
You can copy one of the input examples into the input field and adapt it to the new equation.}
\ \\ \ \\
Finding the solution for each case and checking the case conditions leads to the solution set
\MEquationItem{$\ML$}{\MLParsedQuestion{10}{2,-6}{3}{LVI}} for the equation $|6+3x|=12$.

\begin{MHint}{Solution}
\begin{eqnarray*}
\ML \;=\; \{-6\MElSetSep 2\}
\end{eqnarray*}
\end{MHint}\\
\MInputHint{Mengen können in der Form \texttt{$\lbrace$a;b;c;\ldots$\rbrace$} eingegeben werden.}
\end{MExercise}

You can practise the stepwise solution of absolute value equations within the following exercise. 


\MDirectRouletteExercises{abs_equations.rtex}{VBKM02_ABSEQTRAINING}


\end{MXContent}

\begin{MXContent}{Mixed Equations}{Mixed Equations}{STD}
\MDeclareSiteUXID{VBKM02_GemischteGleichungen}

\begin{MInfo}
If an equation contains both absolute values and other expressions, the case analysis 
has to be done according to the absolute value terms and applied only to these.
\end{MInfo}

Finally, keep in mind to crosscheck the found solution sets with the case conditions.

\begin{MExample}
Solve the equation $|x-1|+x^2=1$. Here, the case analysis is as follows:
\begin{itemize}
\item{For $x\geq 1$, the absolute value bars can be replaced by normal brackets which results
in the quadratic equation $(x-1)+x^2=1$ that is transformed into the equation $x^2+x-2=0$.
Using the $p q$ formula we get the two solutions
\begin{eqnarray*}
x_1 & =& -\Mdfrac12-\sqrt{\Mdfrac94}\;=\; -2 \MDFPSpace, \\
x_2 & =& -\Mdfrac12+\sqrt{\Mdfrac94}\;=\; 1 \MDFPSpace
\end{eqnarray*}
of which only $x_2$ satisfies the case condition.
}
\item{For $x<1$, one obtains the quadratic equation $-(x-1)+x^2=1$ that is 
transformed into the equation $x^2-x=0$ or $x\cdot (x-1)=0$, respectively. The product representation 
indicates the two solutions $x_3=0$ and $x_4=1$. Because of the case condition only 
$x_3=0$ is a solution of the initial equation.
}
\end{itemize}
So, altogether $\ML=\lbrace 0\MElSetSep 1\rbrace$ is the solution set of the 
initial equation.
\end{MExample}

%TODO: Intervalle sind hier noch garnicht erklaert
%TODO: Kodierung von offenen Intervallgrenzen in "\MLIntervalQuestion" (?!)
\begin{MExercise}
Find the solution set of the mixed equation $|x-3|\cdot x=9$.
\begin{MExerciseItems}
\item{If $x$ is in the interval \MLIntervalQuestion{14}{[3,infty)}{5}{GIM1} the  
absolute value term is non-negative.\\ One obtains the quadratic equation 
\MEquationItem{\MLFunctionQuestion{15}{x^2-3*x-9}{5}{x}{5}{GIM2}}{$0$}.\\
The solution set is \MLParsedQuestion{35}{3/2-sqrt(45/4)\MElSetSep 3/2+sqrt(45/4)}{5}{GIM3}.\\
Only the solution \MLParsedQuestion{15}{3/2+sqrt(45/4)}{5}{GIM4} satisfies the case condition.}

\item{If $x$ is in the interval \MLIntervalQuestion{14}{(-infty,3)}{5}{GIM5} the absolute 
value term is negative.\\  One obtains the normalised quadratic equation 
\MEquationItem{\MLFunctionQuestion{15}{x^2-3*x+9}{5}{x}{5}{GIM6}}{$0$}.\\
The solution set is \MLParsedQuestion{30}{}{5}{GIM7}.}
\end{MExerciseItems}
%TODO: Text und Kodierung anpassen
\MInputHint{Enter open intervals in the form $(3;5)$ and closed intervals in the form $[3;5]$. Enter
``infinity'' as text or simply as \texttt{infty}. Do not use the notation $]a;b[$ fir open intervals. 
Sets can be entered by listing the elements in the form $\lbrace 1\MElSetSep 2\MElSetSep 3\rbrace$. For 
the set brackets enter AltGr+7 or AltGr+0, respectively.} 

So, altogether the solution set is \MEquationItem{$\ML$}{\MLParsedQuestion{20}{3/2+sqrt(45/4),3/2+sqrt(45/4)}{5}{GIM8}}.

\begin{MHint}{Solution}
If $x$ is in the interval $\left[3\MIntvlSep \infty\MoIr[\right]$ the absolute value term is non-negative
and one obtains the quadratic equation $x^2-3x-9=0$ with the solution set 
$\ML=\lbrace \Mtfrac32-\sqrt{\Mtfrac{45}{4}};\Mtfrac32+\sqrt{\Mtfrac{45}{4}}\rbrace$. Only the 
larger solution $\Mtfrac32+\sqrt{\Mtfrac{45}{4}}$ satisfies the condition $x\geq 3$. This can 
also be seen without any calculator by estimating $\sqrt{\Mtfrac{45}{4}}\geq\sqrt{\Mtfrac{36}{4}}=3$.
In contrast, if $x$ is in the interval $\MoIl[\left]-\infty\MIntvlSep 3\MoIr[\right]$
the absolute value term is negative. One obtains the normalised quadratic equation $x^2-3x+9=0$.
Because of $\Mtfrac14p^2-q<0$ in the $pq$ formula this equation is unsolvable. Hence, the initial
equation has only one solution $\Mtfrac32+\sqrt{\Mtfrac{45}{4}}$.
\end{MHint}
\end{MExercise}

\begin{MExercise}
Find the solutions of the mixed absolute value equation $3|2x+1|=|x-5|$ by visualising 
the different cases on the number line and finally solving the equation by case analysis. 
First, visualise the case analysis for each absolute value. 

The solution set is \MLParsedQuestion{10}{-8/5,2/7}{5}{PARSEDQUEST1}.

\begin{MHint}{Solution}
Visualising the different cases for the expressions $|2x+1|$ and $|x-5|$ above each other 
indicates all cases to be distinguished:

%%\MGraphics{fallunterscheidung3.png}{scale=1}{Graphische Darstellung der drei Fälle.\MCopyrightLabel{VBKM02_Grafik_Fallunterscheidung3}}
%%\MCopyrightNotice{\MCCLicense}{NONE}{VEMINT}{Im Rahmen des VE\&MINT-Projekts}{VBKM02_Grafik_Fallunterscheidung3}
\begin{center}
\MTikzAuto{%
\begin{tikzpicture}[x=0.35cm, y=1.0cm] 
\fill[color=yellow!20!white] (-7,-1.2) rectangle (15,6.2);
\foreach \sy in {0.0cm,2.5cm,5.0cm} {
\begin{scope}[xshift=0,yshift=\sy]
\draw[-stealth',black] (-6,0) -- (13.5,0);
\foreach \x in {-4, 0, 4, 8, 12}
\draw[shift={(\x,0)},color=black] (0pt,0pt) -- (0pt,3.0pt) (0pt,0pt) node[below=1.0pt] {\small $\x$};
\foreach \x in {-5, -3, -2, -1, 1, 2, 3, 5, 6, 7, 9, 10, 11}
\draw[shift={(\x,0)},color=black] (0pt,0pt) -- (0pt,1.0pt);
\end{scope}
\def\rdst{0.2}
\def\tdst{0.6}
\def\vdst{0.4}
\draw[color=red,line width=2.2pt] (-6.0,\rdst) -- (-0.5,\rdst);
\draw[color=green!60!black,line width=2.2pt] (-0.5,\rdst) -- (5.0,\rdst);
\draw[color=orange,line width=2.2pt] (5.0,\rdst) -- (12.0,\rdst);
\draw[color=red,line width=2.2pt] (-3.25,\tdst) node {Case (1)};
\draw[color=green!60!black,line width=2.2pt] (2.25,\tdst) node {Case (2)};
\draw[color=orange,line width=2.2pt] (8.5,\tdst) node {Case (3)};
\draw[color=blue,line width=0.8pt] (-0.5,\vdst) -- (-0.5,-0.5) node[below=0pt] {$-\frac{1}{2}$} ;
\draw[color=blue,line width=0.8pt] (5.0,\vdst) -- (5.0,-0.5) node[below=0pt] {$5$} ;
\begin{scope}[xshift=0,yshift=2.5cm]
\draw[color=red,line width=2.2pt] (-6.0,\rdst) -- (5.0,\rdst);
\draw[color=orange,line width=2.2pt] (5.0,\rdst) -- (12.0,\rdst);
\draw[color=red,line width=2.2pt] (-0.50,\tdst) node {$x-5<0$};
\draw[color=orange,line width=2.2pt] (8.5,\tdst) node {$x-5>0$};
\draw[color=blue,line width=0.8pt] (5.0,\vdst) -- (5.0,-0.5) node[below=0pt] {$5$} ;
\end{scope}
\begin{scope}[xshift=0,yshift=5cm]
\draw[color=red,line width=2.2pt] (-6.0,\rdst) -- (-0.5,\rdst);
\draw[color=orange,line width=2.2pt] (-0.5,\rdst) -- (12.0,\rdst);
\draw[color=red,line width=2.2pt] (-3.25,\tdst) node {$2x+1<0$};
\draw[color=orange,line width=2.2pt] (5.75,\tdst) node {$2x+1>0$};
\draw[color=blue,line width=0.8pt] (-0.5,\vdst) -- (-0.5,-0.5) node[below=0pt] {$-\frac{1}{2}$} ;
\end{scope}
}
\end{tikzpicture}
}
\par
Illustration of the three cases
\end{center}

According to the figure above the following three cases have to be distinguished:
\begin{itemize}
\item{Case (1): For $x<-\Mtfrac12$ both terms in the absolute value terms are negative.}
\item{Case (2): For $-\Mtfrac12\leq x<5$ the term in the second absolute value term is negative but the 
term in the first one is not.}
\item{Case (3): For $5\leq x$ both terms in the absolute value terms are non-negative.}
\item{Obviously, there is no $x$ for which the first term is negative and the second 
term is non-negative.}
\end{itemize}

So, the solutions can be summarised:
\begin{itemize}
\item{In case (1), both absolute values reverse the sign of the terms:\\ $3|2x+1|=|x-5|\;\Leftrightarrow\;3(-(2x+1)) = -(x-5)$.\\
This equation has the solution $x=-\Mtfrac85$ satisfying the case condition.}
\item{In case (2), only the second absolute value reverses the sign of the term: \\$3|2x+1|=|x-5|\;\Leftrightarrow\;3(2x+1) = -(x-5)$.\\
This equation has the solution $x=\Mtfrac27$ satisfying the case condition.}
\item{In case (3), the absolute value bars in both terms can be omitted (replaced by normal brackets): \\
 $3|2x+1|=|x-5|\;\Leftrightarrow\;3(2x+1) = (x-5)$.\\
This equation has the solution $x=-\Mtfrac85$, but this solution does \textit{not} 
satisfy the case condition. Thus, it will be discarded within its case analysis.}
\end{itemize}
Therefore, the solution set is $\lbrace -\Mtfrac85\MElSetSep \Mtfrac27\rbrace$.
\end{MHint}
\end{MExercise}


\end{MXContent}

\MSubsection{Final Test}
\MLabel{M02_Abschlusstest}

\begin{MTest}{Final Test Modul 2}
\MDeclareSiteUXID{VBKM02_Abschlusstest}

\begin{MExercise}
Find an absolute value term describing the following graph of a function as easy as possible:

%%\MUGraphics{abs2.png}{width=0.5\linewidth}{Funktionsgraph von $f(x)$.}{width:300px}
\begin{center}
\MTikzAuto{%
\begin{tikzpicture}[x=0.8cm, y=1.12cm] 
\draw[black] (-5,0) -- (5,0) (0,-5) -- (0,5);
\foreach \x in {-4, -2, 2, 4}
\draw[shift={(\x,0)},color=black] (0pt,0pt) -- (0pt,-3.0pt) node[below=1.0pt] {\normalsize $\x$};
\foreach \x in {-4.8, -4.4, ..., 5.0}
\draw[shift={(\x,0)},color=black] (0pt,0pt) -- (0pt,-1.5pt);
\foreach \y in {-4, -2, 2, 4}
\draw[shift={(0,\y)},color=black] (0pt,0pt) -- (-3.0pt,0pt) node[left=1.0pt] {\normalsize $\y$};
\foreach \y in {-4.8, -4.4, ..., 5.0}
\draw[shift={(0,\y)},color=black] (0pt,0pt) -- (-1.5pt,0pt);
%%\draw[black] (-0.0pt,-0.5pt) node[anchor=north east] {\small $0$};
\clip(-5.0,-5.0) rectangle (5.0,5.0);
\draw[black, line width=1.0pt,color=black] (-1,6) -- (3,-2) -- (5,2);
\end{tikzpicture}
}
\par
Graph of the function $f(x)$.
\end{center}

Answer: $f(x)=$ \MLSimplifyQuestion{20}{2*abs(x-3)-2}{5}{x}{5}{0}{KA0}\: .
\end{MExercise}

\begin{MExercise}
Solve the following equations:
\begin{MExerciseItems}
\item{$|2x-3|=8$ has the solution set \MLParsedQuestion{16}{11/2,-5/2}{5}{KA1}.}
\item{$|x-2|\cdot x=0$ has the solution set \MLParsedQuestion{16}{0,2}{5}{KA2}.} 
\end{MExerciseItems}
\MInputHint{Enter sets in the form \texttt{$\lbrace$ a;b;c;$\ldots\rbrace$}. 
Enter the empty set as $\lbrace\rbrace$.}
\end{MExercise}

\begin{MExercise}
A camera has a resolution of $6$ megapixels, i.e. -- for convenience -- 
6 million pixels and produces images in format $2:3$. Which size has a 
quadratic pixel on a print-out of format $(60$ cm$) \times (40$ cm$)$? 
Specify the side length of a pixel in millimetre. 

Answer: \MLParsedQuestion{10}{0.2}{5}{VPIXQ}\ \ \ \ (without the unit mm).
\end{MExercise}

\begin{MExercise}
Find the solution set of the mixed equation $|x-1|\cdot (x+1)=3$.\\
Answer: \MEquationItem{$\ML$}{\MLParsedQuestion{15}{2,2}{5}{KAX4}}.
\end{MExercise}


\end{MTest}


\newpage
\MPrintIndex

\end{document}
